Nonparametric selective bootstrap

### 2015-05-27 notes/nonparam.tex
pairs version of maxT

\newcommand{\real}{\mathbb{R}}

@@

For large enough $n$, under our model $M$ suppose that
$$
\gamma(\hat{F}_n) \approx N(\gamma(F), n^{-1} \Sigma(F))
$$
where
$$
\Sigma(F) = E_{F}\left((\gamma(t) - \gamma(F))(\gamma(t) - \gamma(F))^T \right).
$$

@@

If this is our reference distribution, we should base our test on the 
conditional law 
$$n^{1/2}\Sigma(F)^{-1}\gamma(\hat{F}_n) \big \vert n^{1/2} P_{\eta}^{\perp}\Sigma(F)^{-1}\gamma(\hat{F}_n).$$ 
where $P_{\eta}=\eta\eta^{\dagger}$ and $P_{\eta}^{\perp} = I - P_{\eta}$.
This is because
we know that, under $H_0:\eta^T\gamma(F)=0$
this distribution does not depend on $P_{\eta}^{\perp}\gamma(F)$ which
is everything except what we wanted to test.

This 
law has covariance $n^{-1} \Theta(\eta, \Sigma(F))$ where
$$
\Theta(\eta, \Sigma(F)) = \left(\Sigma(F)^{-1} - \Sigma(F)^{-1} P_{\eta}^{\perp} \left(P_{\eta}^{\perp} \Sigma(F)^{-1} P_{\eta}^{\perp} \right)^{\dagger}  P_{\eta}^{\perp} \Sigma(F)^{-1}  \right)
$$
which is an $O(1)$ covariance matrix.
Under $H_0$, this law has mean 0. Under any alternative, its mean grows like $n^{1/2}$.

Under $H_0$
this distribution can be realized by 
$$
L(\eta, \Sigma(F), W) = n^{1/2} \left(\Sigma(F)^{-1} W - \Sigma(F)^{-1} P_{\eta}^{\perp} \left(P_{\eta}^{\perp} 
\Sigma(F)^{-1} P_{\eta}^{\perp} \right)^{\dagger}  P_{\eta}^{\perp} W \right)
$$
where $W \sim N(\mu, n^{-1}\Sigma(F))$, is any normal distribution with
$\eta^T\mu=0$.

In other words, if $W \sim N(\mu, n^{-1}\Sigma(F))$ 
then under $H_0: \eta^T\gamma(F)=0$ we have, asymptotically
$$
 L(\eta, \Sigma(F), \gamma(\hat{F}_n))  \overset{n \rightarrow \infty}{\Rightarrow} L(\eta, \Sigma(F), W) = N(0,  \Theta(\eta, \Sigma(F))).
$$

@@

We apply non-parametric bootstrap to the pairs 
yielding samples
$$
t^{*,b} \in {\cal X}^n, \qquad 1 \leq  b \leq B
$$
and  bootstrapped sufficient statistics
$$
\gamma^{*,b} = \frac{1}{n} \sum_{i=1}^n \gamma(t_i^{*,b}),  \qquad 1 \leq  b \leq B
$$

These are of course just evaluating $\gamma$ on 
IID draws of size $n$ from $\hat{F}_n$. So,
we can write
$\gamma^{*,b} = \gamma(\hat{F}_n^{*,b})$
where $\hat{F}_n^{*,b}$ is the empirical
distribution function of sampling with replacement from the original data.

For large enough $n$, the law of $\gamma(\hat{F}_n^{*,b})$ is going to be converging
to the Gaussian 
$$
\gamma(\hat{F}_n^{*,b}) | \hat{F}_n \sim N(\gamma(\hat{F}_n), n^{-1} \Sigma(\hat{F}_n))
$$
where
$$
\Sigma(\hat{F}_n) = \frac{1}{n} \sum_{i=1}^n \left(\gamma(t_i) - \gamma(\hat{F}_n) \right)\left(\gamma(t_i) - \gamma(\hat{F}_n) \right)^T.
$$

Therefore, our unselective bootstrap should be giving samples from
roughly $ N(\gamma(\hat{F}_n), n^{-1} \Sigma(\hat{F}_n))$. That is,
taking $B$ very large:
$$
\frac{1}{B}\sum_{b=1}^B \eta^T \gamma(\hat{F}^{*,b}_n) \approx \eta^T\gamma(\hat{F}_n).
$$
This suggests we correct the unselective bootstrap by drawing
$$
\gamma(\hat{F}^{*,b}_n) - P_{\eta}\gamma(\hat{F}_n) | \hat{F}_n \approx N(P_{\eta}^{\perp}\gamma(\hat{F}_n) ,n^{-1} \Sigma(\hat{F}_n)).
$$
With this correction, we see that
$$
L(\eta, \Sigma(\hat{F}_n), \gamma(\hat{F}^{*,b}_n) - P_{\eta} \gamma(\hat{F}_n)) \big \vert \hat{F}_n
 \approx N(0 ,n^{-1} \Theta(\eta, \Sigma(\hat{F}_n))).
$$
As $n$ gets larger, $\Sigma(\hat{F}_n)$ will approach $\Sigma(F)$. 
If the RHS above were actually $$N(0 ,n^{-1} \Theta(\eta, \Sigma(F))),$$ then
by our earlier observation at the end of Section \ref{sec:selective} this is sufficient to carry out the selective test exactly.
However, the variance is not exactly correct. In general,
plugging in consistent estimates of variance is OK in selective inference (not sure
if Xiaoying and I put it in the first paper on asymptotics, but she has given me a proof and definitely in the
randomized paper we will have a concrete statement about this). In Rob's example of $n=30$, the sample size might be too small still, but asymptotically
I think it will be fine. Maybe a degrees of freedom correction will help.

The final algorithm is:
we draw nonparametric bootstrap samples
$$
\gamma(\hat{F}^{*,b}_n) - P_{\eta}\gamma(\hat{F}_n)
$$
keeping only those for which $\hat{M}(\gamma(\hat{F}^{*,b}_n) - P_{\eta}\gamma(\hat{F}_n)) = \hat{M}(\gamma(\hat{F}_n))$.

For those that survive, we compute
$$
L(\eta, \Sigma(\hat{F}_n), \gamma(\hat{F}^{*,b}_n) - P_{\eta}\gamma(\hat{F}_n))
$$
and take any test statistic of our choosing.

@@

When there are no nuisance parameters (i.e. $\eta=I$ or we are make other assumptions about $Z(F)$), we do not need the
function $L$. This is the example I tried this morning. In this
case, $\gamma=Z$ and we just draw nonparametric bootstrap samples
$$
Z(\hat{F}^{*,b}_n) - P_{\eta}Z(\hat{F}_n)
$$
keeping the ones that satisfy selection and compute any test statistic of our 
choosing on these bootstrap samples.

>>>>

\newcommand{\real}{\mathbb{R}}
\usepackage{mycommands}

@@

We will follow the convention of \cite{penn_pairs} in stating
a bootstrap CLT as an $m$-of-$n$ bootstrap with $m \gg n$.
For large enough $m$, under our model $M$ suppose that
$$
\gamma(\hat{F}_m) \approx N(\gamma(F), m^{-1} \Sigma(F))
$$
where
$$
\Sigma(F) = \Sigma(F,\gamma) = E_{F}\left((\gamma(t) - \gamma(F))(\gamma(t) - \gamma(F))^T \right).
$$

The bootstrap version of this result will be denoted by
$$
\gamma(\hat{F}^{*,m}_n) \overset{m \to \infty}{\Rightarrow} N(\gamma(\hat{F}_n), m^{-1} \Sigma(\hat{F_n})).
$$
That is, if we take $m \gg n$ samples with replacement from $\hat{F}_n$, 
a CLT holds in $m$. The statement ``the bootstrap works'' can then
be interpreted as 
$$
\gamma(\hat{F}^{*,n}_n) \overset{n \to \infty}{\Rightarrow} N(\gamma(F), n^{-1} \Sigma(F)).
$$

@@

If this is our reference distribution, we should base our test on the 
conditional law 
$$m^{1/2}\Sigma(F)^{-1}\gamma(\hat{F}_m) \big \vert m^{1/2} P_{\eta}^{\perp}\Sigma(F)^{-1}\gamma(\hat{F}_m).$$ 
where $P_{\eta}=\eta\eta^{\dagger}$ and $P_{\eta}^{\perp} = I - P_{\eta}$.
This is because
we know that, under $H_0:\eta^T\gamma(F)=0$
this distribution does not depend on $P_{\eta}^{\perp}\gamma(F)$ which
is everything except what we wanted to test.

This 
law has covariance $m^{-1} \Theta(\eta, \Sigma(F))$ where
$$
\Theta(\eta, \Sigma(F)) = \left(\Sigma(F)^{-1} - \Sigma(F)^{-1} P_{\eta}^{\perp} \left(P_{\eta}^{\perp} \Sigma(F)^{-1} P_{\eta}^{\perp} \right)^{\dagger}  P_{\eta}^{\perp} \Sigma(F)^{-1}  \right)
$$
which is an $O(1)$ covariance matrix.
Under $H_0$, this law has mean 0. Under any alternative, its mean grows like $m^{1/2}$.

Under $H_0$
this distribution can be realized by 
$$
L(\eta, \Sigma(F), W) = m^{1/2} \left(\Sigma(F)^{-1} W - \Sigma(F)^{-1} P_{\eta}^{\perp} \left(P_{\eta}^{\perp} 
\Sigma(F)^{-1} P_{\eta}^{\perp} \right)^{\dagger}  P_{\eta}^{\perp} W \right)
$$
where $W \sim N(\mu, m^{-1}\Sigma(F))$, is any normal distribution with
$\eta^T\mu=0$.

In other words, if $W_m \sim N(\mu, m^{-1}\Sigma(F))$ 
then under $H_0: \eta^T\gamma(F)=0$ we have, asymptotically
$$
 L(\eta, \Sigma(F), \gamma(\hat{F}_m))  \overset{m \rightarrow \infty}{\Rightarrow} L(\eta, \Sigma(F), W_m) = N(0,  \Theta(\eta, \Sigma(F))).
$$

@@

We apply non-parametric bootstrap drawing $m$ samples
yielding 
$$
t^{*,b} \in {\cal X}^m, \qquad 1 \leq  b \leq B
$$
and  bootstrapped sufficient statistics
$$
\gamma^{*,b} = \frac{1}{m} \sum_{i=1}^m \gamma(t_i^{*,b}),  \qquad 1 \leq  b \leq B
$$

These are of course just evaluating $\gamma$ on 
IID draws of size $m$ from $\hat{F}_n$. So,
we can write
$\gamma^{*,b} = \gamma(\hat{F}_n^{*,m})$
where $\hat{F}_n^{*,m}$ is the empirical
distribution function of sampling with replacement from the original data.

For large enough $m$, the law of $\gamma(\hat{F}_n^{*,m})$ is going to be converging
to the Gaussian 
$$
\gamma(\hat{F}_n^{*,m}) | \hat{F}_n \sim N(\gamma(\hat{F}_n), m^{-1} \Sigma(\hat{F}_n)).
$$

Therefore, our unselective bootstrap should be giving samples from
roughly $ N(\gamma(\hat{F}_n), m^{-1} \Sigma(\hat{F}_n))$. That is,
taking $B$ very large:
$$
\frac{1}{B}\sum_{b=1}^B \eta^T \gamma(\hat{F}^{*,m}_n) \approx \eta^T\gamma(\hat{F}_n).
$$
This suggests we correct the unselective bootstrap by drawing
$$
\gamma(\hat{F}^{*,m}_n) - P_{\eta}\gamma(\hat{F}_n) | \hat{F}_n \approx N(P_{\eta}^{\perp}\gamma(\hat{F}_n) ,m^{-1} \Sigma(\hat{F}_n)).
$$
With this correction, we see that
$$
L(\eta, \Sigma(\hat{F}_n), \gamma(\hat{F}^{*,m}_n) - P_{\eta} \gamma(\hat{F}_n)) \big \vert \hat{F}_n
 \approx N(0 ,m^{-1} \Theta(\eta, \Sigma(\hat{F}_n))).
$$

Now, as $n$ gets larger, $\Sigma(\hat{F}_n)$ will approach $\Sigma(F)$. 
If the RHS above were actually $$N(0 ,m^{-1} \Theta(\eta, \Sigma(F))),$$ then
by our earlier observation at the end of Section \ref{sec:selective} this is sufficient to carry out the selective test exactly.
However, the variance is not exactly correct. In general,
plugging in consistent estimates of variance is OK in selective inference (not sure
if Xiaoying and I put it in the first paper on asymptotics, but she has given me a proof and definitely in the
randomized paper we will have a concrete statement about this). In Rob's example of $n=30$, the sample size might be too small still, but asymptotically
I think it will be fine. Maybe a degrees of freedom correction will help.

The final algorithm is:
we draw nonparametric bootstrap samples
$$
\gamma(\hat{F}^{*,m}_n) - P_{\eta}\gamma(\hat{F}_n)
$$
keeping only those for which $\hat{M}(\gamma(\hat{F}^{*,m}_n) - P_{\eta}\gamma(\hat{F}_n)) = \hat{M}(\gamma(\hat{F}_n))$.

For those that survive, we compute
$$
L(\eta, \Sigma(\hat{F}_n), \gamma(\hat{F}^{*,m}_n) - P_{\eta}\gamma(\hat{F}_n))
$$
and take any test statistic of our choosing.

@@

When there are no nuisance parameters (i.e. $\eta=I$ or we are make other assumptions about $Z(F)$), we do not need the
function $L$. This is the example I tried this morning. In this
case, $\gamma=Z$ and we just draw nonparametric bootstrap samples
$$
Z(\hat{F}^{*,m}_n) - P_{\eta}Z(\hat{F}_n)
$$
keeping the ones that satisfy selection and compute any test statistic of our 
choosing on these bootstrap samples.

\section{Pairs bootstrap regression}

A lot of this is based on the excellent survey \cite{penn_pairs}, though 
I want to turn it into selected model language. In this section,
$X$ is a random vector on $\real^p$, not a matrix, and $y$ is a 
random variable, not a vector. 
Given a distribution $F$ on $\real \times \real^p$, there are various
quantities that appear in the ``pairs bootstrap'' for regression.
(We could assume that we have removed a constant mean
from both $X$ and $y$, so that 
$\E_F((y,X)) = 0$ though nothing strictly assumes this unless
we want to talk about ``covariances'' and ``correlations'' instead
 of ``inner productes'' and ``angles''.)

In any case, least squares regression is a statistical functional
$$
\beta(F) = \argmin_{\beta \in \real^P} \E_F((y-X^T\beta)^2)
$$
where $(y,X) \sim F$.
This leads us to the definition of the total error, a random vector 
defined in \cite{penn_pairs}
$$
\delta = \delta(F) = y - X^T\beta,  \qquad (y, X) \sim F.
$$
They further decompose this error into two components
$$
\delta(F) = \epsilon(F) + \zeta(F) = (y - \E_F(y|X)) + (\E_F(y|X) - X^T\beta).
$$
(I don't want to overload $\eta$, so I used $\zeta$ in place of their $\eta$).

Note that we can rewrite the error equation as
\begin{equation}
\label{eq:datagen:full}
y = X^T\beta(F) +\delta(F), \qquad (y, X) \sim F.
\end{equation}
This says that, in order to draw from $F$ it is sufficient to know 
$\beta(F)$ and draw from $(X, \delta(F))$. This formula
has a more familiar ``data-generating'' mechanism to it.

One of the main takeaways of\cite{penn_pairs} is that the $m$-of-$n$ pairs bootstrap distribution of $\beta(\hat{F}_n)$ for any fixed $n$ (i.e. drawing $m \gg n$ samples with replacement
from $\hat{F}_n$ and running OLS on this data set) 
has the following CLT
$$
m^{1/2} \left(\beta(\hat{F}^{*,m}_n) - \beta(\hat{F}_n) \right)
\overset{M\to \infty}{\Rightarrow} N(0, S(\hat{F}_n))
$$
where 
$$
S(F) = (\E_F(XX^T))^{-1} \E_F(\delta(F)XX^T)(\E_F(XX^T))^{-1}
$$
is the sandwich estimator for the law $F$ and 
$\hat{F}^{*,m}_n$ is a sample of size $M$ with replacement from $\hat{F}_n$.

The fact that the pairs bootstrap works as $n \to \infty$ can therefore be summarized as saying that
$$
S(\hat{F}_n) \overset{n \rightarrow \infty}{\to} S(F).
$$

\subsection{Selected model tests}

We are of course interested in selective versions of these statements.
Let $A \subset [p]=\{1, \dots, p\}$ and $I=A^c$ partition
the variables into active and inactive sets.
Then, we can write
\begin{equation}
\label{eq:datagen}
y = X_A\beta_A(F) +\delta_A(F), \qquad (y, X) \sim F
\end{equation}
which makes sense whether or not $\beta(F)[I]=0$ or not. Note that
we should rewrite \eqref{eq:datagen:full} as
\begin{equation}
\label{eq:datagen}
y = X\beta_{[p]}(F) +\delta_{[p]}(F), \qquad (y, X) \sim F
\end{equation}

What happens if $\beta(F)_I=0$? In this case,
$$
X^T\beta_{[p]}(F) = X_A^T\beta(F)_A.
$$
However, it is not immediately clear 
$$
\beta_{[p]}(F)_A=\beta_A(F).
$$
For this to be the case, it would seem to have to be that
$$
\delta_A(F)=\delta_{[p]}(F)
$$
which we have no reason to believe is true in general. In fact, I think
we have
$$
\beta_{[p]}(F)_A = \beta_A(F), \beta_{[p]}(F)_I=0 \iff \delta_A(F)=\delta_{[p]}(F).
$$

\subsection{Analog of max-$T$}

In the fixed design case, our max-$T$ test looks at the law of
$yX_I$ adjusted for those variables in $A$. In the
fixed design case, these adjustments are done
in matricial form which we are not using in this section.
Morally, though, they use the residuals after having adjusted for $A$
and see if there are any other important correlations remaining.

A natural analog
in the pairs bootstrap model is to look at the law of
$$
\begin{aligned}
\delta_A(F) X_I &= (y-X_A^T\beta_A(F)) X_I \\
&= yX_I - X_I \cdot X_A^T\beta_A(F)
\end{aligned}
$$
and perhaps test $H_0:\E_F(\delta_A(F)X_I)=0$.
From our earlier discussion we see that
$$
\begin{aligned}
\delta_A(F)X_I &= (X\beta_{[p]}(F) + \delta_{[p]}(F) - X_A\beta_A(F))X_I \\
&= X_I(X_I^T\beta_{[p]}(F)_I + X_A^T(\beta_{[p]}(F)_A - \beta_A(F))  + \delta_{[p]}(F))
\end{aligned}
$$
and
$$
\E_F(\delta_A(F)X_I) = \E_F(X_IX_I^T)\beta_{[p]}(F)_I + \E_F(X_IX_A^T)(\beta_{[p]}(F)_A - \beta_A(F)).
$$
Therefore, testing $H_0:\E_F(\delta_A(F)X_I)=0$ is close to
to testing $\tilde{H}_0:\beta_{[p]}(F)_I=0, \beta_{[p]}(F)_A-\beta_A(F)=0$ but not
quite the same. However, if we reject $H_0$ then we know one of the
two conditions in $\tilde{H}_0$ holds.
The complication here is that $\beta_A(F)$ appears in the law
we are interested in, i.e. the law of $\delta_A(F)X_I$.

Even in the fixed design case, we use the empirically adjusted residuals
instead of the actual residuals. We might as well do the same here.
Define an empirical version as the statistical functional
$$
\begin{aligned}
\tilde{\gamma}_{A}(F) &= \E_F(yX_I - \E_F(X_I X_A^T) \E_F(X_AX_A^T)^{-1} yX_A) \\
&= \E_F(X_I \delta_A(F) + X_IX_A^T\beta_A(F) - \E_F(X_I X_A^T) \E_F(X_AX_A^T)^{-1} X_A(X_A^T\beta_A(F) + \delta_A(F))) \\
&= \E_F(X_I \delta_A(F)).
\end{aligned}
$$
This has the same expected value as 
the law we were interested in and it also has an $m$-of-$n$ bootstrap CLT as well.

If $\mathbf{y}, \mathbf{X}$ is an IID sample of size $n$ from $F$ then
$$
\begin{aligned}
\tilde{\gamma}_A(\hat{F}_n) &= \frac{1}{n} \left(\mathbf{X}_I^T\mathbf{y} - \left(\mathbf{X}_I^T\mathbf{X}_A \right) \left(\mathbf{X}_A^T\mathbf{X}_A \right)^{-1} \mathbf{X}_A^T\mathbf{y} \right) \\
 &= \frac{1}{n}\mathbf{X}_I^T\mathbf{\delta_A(y)} \\
\end{aligned}
$$
where
$$
\mathbf{\delta_A(y)} = \mathbf{y} - \mathbf{X}_A \left(\mathbf{X}_A^T\mathbf{X}_A \right)^{-1} \mathbf{X}_A^T\mathbf{y}
$$
are the empirical OLS residuals.

It would seem that we might have
run OLS for each bootstrap sample which might be expensive. 
However, following \cite{penn_pairs} it is generally OK to replace
matrices of cross-products in expressions. It is OK in the sense
that the bootstrap CLT is generally preserved. For instance, the
bootstrap distribution of $\beta_{[p]}(\hat{F}_n)$ is essentially the same as
that of the functional
$$
\E_F(XX^T)^{-1} yX
$$
which is not a statistical functional per se as it involves $\E_F(XX^T)$.

With this in mind, we might consider the functional 
$$
\begin{aligned}
\gamma_A(\hat{F}_n) &= \frac{1}{n} \left(\mathbf{X}_I^T\mathbf{y} - \E_F(X_IX_A^T) \E_F(X_AX_A^T)^{-1} \mathbf{X}_A^T\mathbf{y} \right)
\end{aligned}
$$
which satisfies
$$
\E_F(\gamma_A(\hat{F}_n)) = \E_F(\tilde{\gamma}_A(\hat{F}_n))
$$
and
$$
n\text{Var}_F(\gamma_A(\hat{F}_n)) \approx n \text{Var}_F(\tilde{\gamma}_A(\hat{F}_n)).
$$
This last assertion about the variance depends on a similar argument
as in the pairs bootstrap which replaces the empirical cross-products
$n^{-1}\mathbf{X}_I^T\mathbf{X}_A$ and $n^{-1}\mathbf{X}_A^T\mathbf{X}_A$
with their expectation at the cost of an error of size $n^{-1/2}$ that does not affect the limiting variance.

One advantage of $\gamma_A$ over $\tilde{\gamma}_A$ is that one does 
not have to solve a least squares problem for every bootstrap sample as
the bootstrap version of $\gamma_A$ would be
$$
\gamma_A(\hat{F}^{*,m}_n) &= \E_{\hat{F}^{*,m}_n}(yX_I) - 
\E_{\hat{F}_n}(X_IX_A^T) (\E_{\hat{F}_n}(X_AX_A^T))^{-1} \E_{\hat{F}^{*,m}_n}(yX_A).
$$

This is a statistic and it has an $m$-of-$n$  bootstrap CLT centered at
$\tilde{\gamma}_A(\hat{F}_n)$ with approximately correct covariance.

\subsection{Sampling correction}

When drawing bootstrap samples we want them to be such that the expected
value of our statistic is 0 under $H_0$. Our bootstrap samples
draw samples of $(yX, XX^T)$ and our selection is a function of these.
To correct, we actually only need to modify the $yX$ part of the sample.
Define 
$$
L_A(F) = \begin{pmatrix} I & \E_{F}(X_IX_A^T) (\E_{F}(X_AX_A^T))^{-1} \end{pmatrix} \in \real^{I \times [p]}
$$
so that
$$
\gamma_A(\hat{F}_n) =  L_A(F)\E_{\hat{F}_n}(yX).
$$

Let $\hat{G}^{*,m}_n$ to be the law of 
$\hat{F}^{*,m}_n$ shifted by
$$
\Delta_A(\hat{F}_n) = L_A(\hat{F}_n)^{\dagger}L_A(\hat{F}_n) \E_{\hat{F}_n}(yX)
$$
so that
$$
\E_{\hat{G}^{*,m}_n}(\gamma_A(\hat{G}^{*,m}_n))= 0.
$$

Then, $\gamma_A(\hat{G}^{*,m}_n)$ satisfies an $m$-of-$n$ bootstrap CLT
with the proper mean under $H_0$ and asymptotically (in $n$) correct variance.

\bibliographystyle{alpha}
\bibliography{nonparam}
